---
title: "Exploration analyses multivariées (chap 2 de Boucard, Gillet & Legendre 2011)"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 6
---

```{r setup, include=FALSE, message=F, warning=F}
knitr::opts_chunk$set(echo = T)
```

## On va loader les packages

```{r, message=FALSE}
require(ade4)
require(vegan)
require(gclus)
require(ape)
```

## On va loader les fonctions utilisées par Legendre

```{r, message=FALSE}
download.file("https://raw.githubusercontent.com/JoeyBernhardt/NumericalEcology/master/panelutils.R", destfile=paste(getwd(), "/panelutils.R", sep=""))
source("panelutils.R")
```

## On va loader les données directement du site 

```{r, message=FALSE}

download.file("https://raw.githubusercontent.com/JoeyBernhardt/NumericalEcology/master/NEwR%20data/DoubsEnv.csv", destfile=paste(getwd(), "/DoubsEnv.csv", sep=""))
download.file("https://raw.githubusercontent.com/JoeyBernhardt/NumericalEcology/master/NEwR%20data/DoubsSpe.csv", destfile=paste(getwd(), "/DoubsSpe.csv", sep=""))
download.file("https://raw.githubusercontent.com/JoeyBernhardt/NumericalEcology/master/NEwR%20data/DoubsSpa.csv", destfile=paste(getwd(), "/DoubsSpa.csv", sep=""))

# Species abundance data frame
spe <- read.csv("DoubsSpe.csv", row.names=1)
# Environmental data frame
env <- read.csv("DoubsEnv.csv", row.names=1)
# Spatial data frame
spa <- read.csv("DoubsSpa.csv", row.names=1)
```

> IMPORTANT NOTE: Legendre extracted these data from the "doubs" dataset in package ade4 version 1.4-10, and we restored the original units of the environmental variables. Names of data frames and variables have changed in the package ade4 since the publication of the book. Moreover, a mistake in the environmental dataset (as compared to the original publication by J. Verneaux) has not yet been corrected in the data currently available from ade4 version 1.5-0 (doubs$env[7,1] = 368 and not 268). Therefore, we assume in our R scripts that you use the CSV files provided here and not the original dataset from ade4!

## Distribution et abondances

```{r}
# Minimum and maximum of abundance values in the whole data set
range(spe)

# Count cases for each abundance class
(ab <- table(unlist(spe)))

# Barplot of the distribution, all species confounded
barplot(ab, las=1, xlab="Abundance class", ylab="Frequency", col=gray(5:0/5))

# Number of absences
sum(spe==0)

# Proportion of zeros in the community data set
sum(spe==0)/(nrow(spe)*ncol(spe))
```

#### Carte de la localisation des sites

```{r}
# Geographic coordinates x and y from the spa data frame
plot(spa, asp=1, type="n", main="Site Locations",
	xlab="x coordinate (km)", ylab="y coordinate (km)")
# Add a blue line connecting the sites (Doubs river)
lines(spa, col="light blue")
# Add site labels
text(spa, row.names(spa), cex=0.8, col="red")
# Add text blocks
text(50, 10, "Upstream", cex=1.2, col="red")
text(30, 120, "Downstream", cex=1.2, col="red")
```

#### Carte de certaines espèces de poissons

```{r}
par(mfrow=c(2,2), mar=c(4,4,2,1))
# Plot four species
plot(spa, asp=1, col="brown", cex=spe$TRU, main="Brown trout", 
	xlab="x coordinate (km)", ylab="y coordinate (km)")
lines(spa, col="light blue")
plot(spa, asp=1, col="brown", cex=spe$OMB, main="Grayling", 
	xlab="x coordinate (km)", ylab="y coordinate (km)")
lines(spa, col="light blue")
plot(spa, asp=1, col="brown", cex=spe$BAR, main="Barbel", 
	xlab="x coordinate (km)", ylab="y coordinate (km)")
lines(spa, col="light blue")
plot(spa, asp=1, col="brown", cex=spe$BCO, main="Common bream", 
	xlab="x coordinate (km)", ylab="y coordinate (km)")
lines(spa, col="light blue")
```

#### Comparer l'occurrence des espèces, le nombre de sites où chaque espèce est présente

```{r}
# To sum by columns, the second argument of apply(), MARGIN, is set to 2
spe.pres <- apply(spe > 0, 2, sum)
# Sort the results in increasing order
sort(spe.pres)
# Compute percentage frequencies
spe.relf <- 100*spe.pres/nrow(spe)
# Round the sorted output to 1 digit
round(sort(spe.relf), 1)

par(mfrow=c(1,2))
hist(spe.pres, main="Species Occurrences", right=FALSE, las=1, 
	xlab="Number of occurrences", ylab="Number of species", 
	breaks=seq(0,30,by=5), col="bisque")
hist(spe.relf, main="Species Relative Frequencies", right=FALSE, las=1,
	xlab="Frequency of occurrences (%)", ylab="Number of species",
		breaks=seq(0, 100, by=10), col="bisque")
```

#### Comparer la richess entre les sites - nombre d'espèces par site

```{r}
# To sum by rows, the second argument of apply(), MARGIN, is set to 1
sit.pres <- apply(spe > 0, 1, sum)
# Sort the results in increasing order
sort(sit.pres)

par(mfrow=c(1,2))
# Plot species richness vs. position of the sites along the river
plot(sit.pres,type="s", las=1, col="gray",
	main="Species Richness vs. \n Upstream-Downstream Gradient",
	xlab="Positions of sites along the river", ylab="Species richness")
text(sit.pres, row.names(spe), cex=.8, col="red")
# Use geographic coordinates to plot a bubble map
plot(spa, asp=1, main="Map of Species Richness", pch=21, col="white", 
	bg="brown", cex=5*sit.pres/max(sit.pres), xlab="x coordinate (km)", 
	ylab="y coordinate (km)")
lines(spa, col="light blue")
```

> On voit qu'il n'y a aucun poisson de récolté dans le site 8!

#### Calculer des indices de diversité des communautés de poissons

```{r}
# Get help on the diversity() function
?diversity

N0 <- rowSums(spe > 0)         # Species richness
H <- diversity(spe)            # Shannon entropy
N1 <- exp(H)                   # Shannon diversity (number of abundant species)
N2 <- diversity(spe, "inv")    # Simpson diversity (number of dominant species)
J <- H/log(N0)                 # Pielou evenness
E10 <- N1/N0                   # Shannon evenness (Hill's ratio)
E20 <- N2/N0                   # Simpson evenness (Hill's ratio)
(div <- data.frame(N0, H, N1, N2, E10, E20, J))
```

### Transformation et standardization des données d'abondance d'espèces

#### Simplement en 0-1

```{r}
# Get help on the decostand() function
?decostand

## Simple transformations

# Partial view of the raw data (abundance codes)
spe[1:5, 2:4]
# Transform abundances to presence-absence (1-0)
spe.pa <- decostand(spe, method="pa")
spe.pa[1:5, 2:4]
```

#### Standardisation par colonne (std selon l'abondance de chaque expèce)

```{r}

# Scale abundances by dividing them by the maximum value for each species
# Note: MARGIN=2 (column, default value) for this method
spe.scal <- decostand(spe, "max")
spe.scal[1:5,2:4]
# Display the maximum by column
apply(spe.scal, 2, max)

# Scale abundances by dividing them by the species totals
# (relative abundance by species)
# Note: MARGIN=2 for this method
spe.relsp <- decostand(spe, "total", MARGIN=2)
spe.relsp[1:5,2:4]
# Display the sum by column
apply(spe.relsp, 2, sum)
```

#### Standardisation par ligne (std selon les sites)

```{r}

# Scale abundances by dividing them by the site totals
# (relative abundance by site)
# Note: MARGIN=1 (default value) for this method
spe.rel <- decostand(spe, "total")
spe.rel[1:5,2:4]
# Display the sum of row vectors to determine if the scaling worked properly
apply(spe.rel, 1, sum)

# Give a length of 1 to each row vector (Euclidean norm)
spe.norm <- decostand(spe, "normalize")
spe.norm[1:5,2:4]
# Verify the norm of row vectors
norm <- function(x) sqrt(x%*%x)
apply(spe.norm, 1, norm)

# Compute square root of relative abundances by site
spe.hel <- decostand(spe, "hellinger")
spe.hel[1:5,2:4]
# Check the norm of row vectors
apply(spe.hel, 1, norm)
```

#### Double standardisation par colonne et ligne (prend en compte la variabilité entre les sites et les espèces)

```{r}

# Chi-square transformation
spe.chi <- decostand(spe, "chi.square")
spe.chi[1:5,2:4]
# Check what happened to site 8 where no species was found
spe.chi[7:9,]

# Wisconsin standardization
# Abundances are first ranged by species maxima and then by site totals
spe.wis <- wisconsin(spe)
spe.wis[1:5,2:4]
```

#### Comparaison des standardisations

```{r}
# Boxplots of transformed abundances of a common species (stone loach)
par(mfrow=c(2,2))
boxplot(spe$LOC, sqrt(spe$LOC), log1p(spe$LOC),
	las=1, main="Simple transformation",
	names=c("raw data", "sqrt", "log"), col="bisque")
boxplot(spe.scal$LOC, spe.relsp$LOC,
	las=1, main="Standardization by species",
	names=c("max", "total"), col="lightgreen")
boxplot(spe.hel$LOC, spe.rel$LOC, spe.norm$LOC,
	las=1, main="Standardization by sites",
	names=c("Hellinger", "total", "norm"), col="lightblue")
boxplot(spe.chi$LOC, spe.wis$LOC,
	las=1, main="Double standardization",
	names=c("Chi-square", "Wisconsin"), col="orange")
```

#### Plotter les transformations

```{r}
# Plot profiles along the upstream-downstream gradient
par(mfrow=c(2,2))
plot(env$das, spe$TRU, type="l", col=4, main="Raw data",
	xlab="Distance from the source [km]", ylab="Raw abundance code")
lines(env$das, spe$OMB, col=3)
lines(env$das, spe$BAR, col="orange")
lines(env$das, spe$BCO, col=2)
lines(env$das, spe$LOC, col=1, lty="dotted")

plot(env$das, spe.scal$TRU, type="l", col=4, main="Species profiles (max)",
	xlab="Distance from the source [km]", ylab="Standardized abundance")
lines(env$das, spe.scal$OMB, col=3)
lines(env$das, spe.scal$BAR, col="orange")
lines(env$das, spe.scal$BCO, col=2)
lines(env$das, spe.scal$LOC, col=1, lty="dotted")

plot(env$das, spe.hel$TRU, type="l", col=4,
	main="Site profiles (Hellinger)",
	xlab="Distance from the source [km]", ylab="Standardized abundance")
lines(env$das, spe.hel$OMB, col=3)
lines(env$das, spe.hel$BAR, col="orange")
lines(env$das, spe.hel$BCO, col=2)
lines(env$das, spe.hel$LOC, col=1, lty="dotted")

plot(env$das, spe.chi$TRU, type="l", col=4,
	main="Double profiles (Chi-square)",
	xlab="Distance from the source [km]", ylab="Standardized abundance")
lines(env$das, spe.chi$OMB, col=3)
lines(env$das, spe.chi$BAR, col="orange")
lines(env$das, spe.chi$BCO, col=2)
lines(env$das, spe.chi$LOC, col=1, lty="dotted")
legend("topright", c("Brown trout", "Grayling", "Barbel", "Common bream",
	"Stone loach"), col=c(4,3,"orange",2,1), lty=c(rep(1,4),3), cex=0.5)
```

## Les données environnementales

```{r}
par(mfrow=c(2,2), mar=c(4,4,2,1))
plot(spa, asp=1, main="Altitude", pch=21, col="white", bg="red",
	cex=5*env$alt/max(env$alt), xlab="x", ylab="y")
lines(spa, col="light blue")
plot(spa, asp=1, main="Discharge", pch=21, col="white", bg="blue",
	cex=5*env$deb/max(env$deb), xlab="x", ylab="y")
lines(spa, col="light blue")
plot(spa, asp=1, main="Oxygen", pch=21, col="white", bg="green3",
	cex=5*env$oxy/max(env$oxy), xlab="x", ylab="y")
lines(spa, col="light blue")
plot(spa, asp=1, main="Nitrate", pch=21, col="white", bg="brown",
	cex=5*env$nit/max(env$nit), xlab="x", ylab="y")
lines(spa, col="light blue")
```

#### Variables environnementales en fonction de la distance de la source

```{r}
par(mfrow=c(2,2))
plot(env$das, env$alt, type="l", xlab="Distance from the source (km)", 
	ylab="Altitude (m)", col="red", main="Altitude")
plot(env$das, env$deb, type="l", xlab="Distance from the source (km)", 
	ylab="Discharge (m3/s)", col="blue", main="Discharge")
plot(env$das, env$oxy, type="l", xlab="Distance from the source (km)", 
	ylab="Oxygen (mg/L)", col="green3", main="Oxygen")
plot(env$das, env$nit, type="l", xlab="Distance from the source (km)", 
	ylab="Nitrate (mg/L)", col="brown", main="Nitrate")
```

#### Correlélgramme de toutes les variables environnementales

```{r}
pairs(env, panel=panel.smooth, diag.panel=panel.hist,
	main="Bivariate Plots with Histograms and Smooth Curves", gap=0.5, cex.labels = 1)
```

#### Transformation des variables environnementales

```{r}
range(env$pen)

# Log-transformation of the slope variable (y = ln(x))
# Compare histograms and boxplots of raw and transformed values
par(mfrow=c(2,2))
hist(env$pen, col="bisque", right=FALSE)
hist(log(env$pen), col="light green", right=F, main="Histogram of ln(env$pen)")
boxplot(env$pen, col="bisque", main="Boxplot of env$pen", ylab="env$pen")
boxplot(log(env$pen), col="light green", main="Boxplot of ln(env$pen)",
	ylab="log(env$pen)")


# Standardization of all environmental variables
# **********************************************

# Center and scale = standardize variables (z-scores)
env.z <- decostand(env, "standardize")
apply(env.z, 2, mean)	# means = 0
apply(env.z, 2, sd)		# standard deviations = 1

# Same standardization using the scale() function (which returns a matrix)
env.z <- as.data.frame(scale(env))
```

